Two applications of logic programming to Coq

### 2020-11-06 papers/types2020/sec3.tex
some more metanotes

\subsection{Proof for the Horn fragment}
\label{ssec:focused}

@@

\begin{figure}[t]
\[
\infer{\XXi_1\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A : T)\in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  s : A}{{E[] \vdash_{CIC} s : A}}% The order should be s : A, not A : s ?
  \qquad
  \infer{\bc{\XXi}{}{[]} A A}{\initLExpert{\XXi}}
  \qquad
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
        {\XXi_1\vdash t : B\qquad
         \bc{\XXi_2}{}l {D[t/x]} A\qquad
         \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
\]
\caption{Specification of the checking of a proof certificate and the
  synthesis of a dependently typed $\lambda$-term.}
\label{fig:augmented}
\end{figure}

@@

Figure~\ref{fig:augmented} contains these proof rules and extends them
in the following fours ways so that it more appropriately fits into
the Coq proof system.
\begin{enumerate}
  \item Instead of having separate connectives for $\forall$ and
    $\supset$, we have the dependent product connective $(x:A)D$. 
  \item We have incorporated both \emph{proof certificates} (using the
    schematic variable $\XXi$) along with expert predicates
    (predicates with the $e$ subscript).
  \item The inference rules are also annotated by terms structures that
    can be give directly to the Coq kernel for checking.
  \item We have added various premises which are responsible to
    pulling information across the Elpi API with Coq.
\end{enumerate}
The proof system described in Figure~\ref{fig:kernel} corresponds to
the Calculus of Inductive Constructions in which all inductive
definitions are limited to the Horn clause.
%
This proof system is inspired by the calculus for proof search in
Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similar to that calculus, we have a
term language that includes terms and lists of terms, and two distinct
typing judgments for the two categories.
%
In our case, the \lP goal formula \lsti{(check Xi (go B T))}
corresponds to $\XXi\vdash B \colon T$ and the formula
%
\lsti{(check Xi (bc D B L))} corresponds to $\XXi\Downarrow L\colon
D\vdash B$.
%
This style of encode coincides with the idea behind the \emph{spine
  calculus}~\cite{Cervesato97tr}.  The main novelty of our proof
system here is that proof terms and proof certificates are used
simultaneously in all inference rules.

The proof system is parameterized by Coq's global environment $E$, which
here can be thought to consist of a set of inductive definitions,
which following Coq's reference manual we denote with
$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  One  rule delegates
to Coq's type checking enforcing the well-sortedness of such types. We
do not have a \emph{local context} since we are only dealing with types
that correspond to Horn clauses, and atomic types are inductively
defined. In fact, we do not have a $\forall$ rule on the right,
although the proof theory would gladly allow it. This means, at the
term level, that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.

The \emph{unfold} rule is the equivalent in our simplified setting of
a \emph{decide} rule in an explicitly focused system
\cite{liang09tcs}.  From a logic programming viewpoint, given an atom,
the rule selects a clause on which to back-chain on: we look-up the
constructors from an \lsti{Inductive} definition from the global
environment, one that matches the atom we wish to backchain on and then
call the latter judgment. The rules for backchaining are standard,
corresponding to the \emph{init} and the left introduction rules for
$\forall$ and $\supset$.

@@

\subsection{Proof certificate checking}

Figure~\ref{fig:kernel} contains the Elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: $\XXi\vdash t \colon A$ corresponds to
\lsti{check $\XXi$ (go A T)} and $\XXi \Downarrow  l \colon D\vdash A$ corresponds to
\lsti{check $\XXi$ (bc D A L)}.

@@

Note that \lsti{prod} encodes dependent products taking a name for pretty
printing, a term and a \lP function from terms to terms: namely $\forall x :
A, B$ is encoded by \texttt{prod ``x'' A (x\textbackslash{} B x)}; when, in the
implementation of the product-left rule, we apply $B$ to the variable
\lsti{Tm}, we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in LJT and PTSC's tradition, the more since proof search in
our application is cut-free.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.
The \lsti{unfold_expert} predicate, then, has the role of selecting which
constructor to introduce for the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase on it.

In Figure~\ref{fig:augmented}  each
inference rule is augmented with an additional premise involving an
\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
These notions are taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
\todo{MM: Talk about randomized FPC?}
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:augmented} and~\ref{fig:kernel}:
that is, the exact set of constructors for the  type of certificates as
well as the specification of the expert predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

@@

\todo{We could move here from elab the paragraph ``It is also possible to pair two different proof certificates, defined
by two different FPC definitions,'' and/or reduce redundancy}
Further, if we view a particular FPC as a means of restricting proofs, it is
possible to build an FPC that \emph{restricts} proofs satisfying two FPCs
simultaneously.
%
% In particular, Figure~\ref{fig:pairing} defines an FPC based on the
% (infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
% \lsti{cert}.
%
The pairing experts for the certificate \lsti{Cert1 <c> Cert2} simply
request that the corresponding experts also succeed for both
\lsti{Cert1} and \lsti{Cert2} 
%
Thus, the query \lsti{check ((qheight 4) <c> (qsize 10)) B}
will succeed if there is a proof of \lsti{B} that has a height less
than or equal to 4 while also being of size less than or equal to 10.

@@

\subsection{A prolog-like tactic}

@@

The glue code between Coq-Elpi and the implementation of our calculus is
straightforward: the goal is given to us as a quadruple of a context, an
evar, a type and a list of extra information. In addition, the certificate is
supplied by the user when calling the tactic (it simply is an integer in this
case). We just need to call \lsti{check} on the goal's type, together with
the certificate, in order to obtain a reconstructed proof term. We avoid
calling the reconstruction directly on the evar because Coq-Elpi ensures that
evars manipulated by \lP are well-typed at all times, and we cannot guarantee
that since we work with partially reconstructed term; getting around this is
simply a matter of assigning the reconstructed term to the evar right
afterwards.
The following example shows how we can use the above tactic to do
FPC-driven logic programming in Coq and return a Coq proof-term:
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog height 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
\end{lstlisting}

@@

If we do not restrict our attention to only Horn clauses then there is
no problem will presenting a focused proof system for all of
first-order intuitionistic logic using the connectives $\forall$ and
$\supset$.  The following rules extend those given earlier.
\[
  \infer[decide]{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[init]{\Pscr\Downarrow A\vdash A}{}
\]
\[
  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
        {\Pscr\Downarrow D[t/x]\vdash A}
  \qquad
  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
        {\Pscr\vdash B\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[\supset R]{\Pscr\vdash B\supset D}{\Pscr, B\vdash D}
  \qquad
  \infer[\forall R]{\Pscr\vdash \forall x.D}{\Pscr\vdash D[y/x]}
\]
As before, the syntactic variable $A$ denotes atomic formulas and, as
usual, the $\forall R$ rule has the restriction that $y$ is not free in
its conclusion.

@@

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog CIC init lp
% LocalWords:  Andreoli's basicstyle lprolog sigs Coq's sortedness Ev
% LocalWords:  backchain gref indt indc coq dest unfoldings str fpc
% LocalWords:  evar nat forall ys elpi Qed lc args ljf

>>>>

\subsection{Proofs for the Horn fragment}
\label{ssec:focused}

@@

\begin{figure}[t]
\[
\infer{\XXi_1\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A : T)\in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  A : s}{{E[] \vdash_{CIC} A : s}}% The order should be s : A, not A : s ?
  %% we are checking nat : Set, for example, so A : s --AM
  \qquad
  \infer{\bc{\XXi}{}{[]} A A}{\initLExpert{\XXi}}
  \qquad
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
        {\XXi_1\vdash t : B\qquad
         \bc{\XXi_2}{}l {D[t/x]} A\qquad
         \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
\]
\caption{Specification of the checking of a proof certificate and the
  synthesis of a dependently typed $\lambda$-term.}
\label{fig:augmented}
\end{figure}

@@

Figure~\ref{fig:augmented} contains these proof rules and extends them
in the following fours ways so that it more appropriately fits into
the Coq proof system.
\begin{enumerate}
  \item Instead of having separate connectives for $\forall$ and
    $\supset$, we have the dependent product connective $(x:A)D$. 
  \item We have incorporated both \emph{proof certificates}~\cite{chihani17jar} (using the
    schematic variable $\XXi$) along with expert predicates
    (predicates with the $e$ subscript). We explain this in Section~\ref{ssec:cert}.
  \item The inference rules are also annotated by terms structures that
    can be give directly to the Coq kernel for checking.
  \item We have added various premises which are responsible to
    pulling information across the Elpi API with Coq.
\end{enumerate}
The proof system described in Figure~\ref{fig:augmented} (and
implemented in Figure~\ref{fig:kernel}) corresponds to the Calculus
of Inductive Constructions in which all inductive definitions are
limited to the Horn clause.
%
This proof system is inspired by the calculus for proof search in
Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similar to that calculus, we have a
term language that includes terms and lists of terms, and two distinct
typing judgments for the two categories.
\begin{metanote}
  This is repeated in 3.2 and isn't too early to talk about code here? -am
\end{metanote}
%
In our case, the \lP goal formula \lsti{(check Xi (go B T))}
corresponds to $\XXi\vdash B \colon T$ and the formula
%
\lsti{(check Xi (bc D B L))} corresponds to $\XXi\Downarrow L\colon
D\vdash B$.
%
This style of  proof terms coincides with the idea behind the \emph{spine
  calculus}~\cite{Cervesato97tr}.  The main novelty of our proof
system here is that proof terms and proof certificates are used
simultaneously in all inference rules.

The proof system is parameterized by Coq's global environment $E$, which
here we take as a set of inductive definitions,
which, following Coq's reference manual, we denote with
$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  One  rule delegates
to Coq's type checking the enforcement of the well-sortedness of such types. We
do not have a \emph{local context} since we are only dealing with types
that correspond to Horn clauses, and atomic types are inductively
defined. In fact, we do not have a $\forall$ rule on the right,
although the proof theory would gladly allow it. This means at the
term level that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.
\begin{metanote}
  Should we call ``unfold'' \emph{decide} (left) and be done with it? AM
\end{metanote}
The \emph{unfold} rule is the equivalent in our simplified setting of
a \emph{decide} rule in an explicitly focused system
\cite{liang09tcs}.  As in the previous proof system for Horn logic,  given an atom,
the rule selects a clause on which to back-chain on: we lookup the
constructors from an \lsti{Inductive} definition from the global
environment, one that matches the atom we wish to backchain on and then
call the latter judgment. The rules for backchaining are standard,
corresponding to the \emph{init} and the (conflation of the) left introduction rules for
$\forall$ and $\supset$.

@@

\subsection{Proof certificate checking}
\label{ssec:cert}
Figure~\ref{fig:kernel} contains the Elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: $\XXi\vdash t \colon A$ corresponds to
\lsti{check $\XXi$ (go A T)} and $\XXi \Downarrow  l \colon D\vdash A$ corresponds to
\lsti{check $\XXi$ (bc D A L)}.

@@

Note that \lsti{prod} encodes dependent products taking a name for pretty
printing, a term and a \lP function from terms to terms: namely $(x :
B) D$ is encoded by \texttt{prod ``x'' B (x\textbackslash{} D x)}; when, in the
implementation of the product-left rule, we apply $D$ to the logic variable
\lsti{Tm}, we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in LJT and PTSC's tradition, the more since proof search in
our application is cut-free.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.
The \lsti{unfoldE} predicate,  has, among others, the role of selecting which
constructor to introduce for the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase on it.

\begin{metanote}
  I still don't like this transition -am
\end{metanote}
In Figure~\ref{fig:augmented}  each
inference rule is augmented with an additional premise involving an
\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
These notions are taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
%%\todo{MM: Talk about randomized FPC? They mentioned afterwards -am}
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:augmented} and~\ref{fig:kernel}:
that is, the exact set of constructors for the  type of certificates as
well as the specification of the expert predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) A)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

@@

\todo{We could move here from elab the paragraph ``It is also possible to pair two different proof certificates, defined
by two different FPC definitions,'' and/or reduce redundancy}
Further, if we view a particular FPC as a means of restricting proofs, it is
possible to build an FPC that \emph{restricts} proofs satisfying two FPCs
simultaneously.
%
% In particular, Figure~\ref{fig:pairing} defines an FPC based on the
% (infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
% \lsti{cert}.
%
The pairing experts for the certificate \lsti{Cert1 <c> Cert2} simply
request that the corresponding experts also succeed for both
\lsti{Cert1} and \lsti{Cert2} 
%
Thus, the query \lsti{check ((qheight 4) <c> (qsize 10)) A}
will succeed if there is a proof of \lsti{A} that has a height less
than or equal to 4 while also being of size less than or equal to 10.

@@

\subsection{A Prolog-like tactic}

@@

The glue code between Coq-Elpi and the implementation of our calculus
is straightforward: the goal is given to us as a quadruple of a (here
inactive) context, an \emph{evar}, a type (goal) and a list of extra
information. In addition, we supply the certificate  when calling the tactic (it simply is an integer and a string to
identify the ``resource'' FPC we will use in this case). We just need
to call \lsti{check} on the goal, together with the
certificate, in order to obtain a reconstructed proof term. We do not
call the reconstruction directly on the evar because Coq-Elpi
ensures that evars manipulated by \lP are well-typed at all times; since
we cannot guarantee that, as we work with partially reconstructed
term, we  get  around this by an explicit unification.

The following
example shows how we can use the above tactic to do FPC-driven logic
programming in Coq and return a Coq proof-term:
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog height 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
\end{lstlisting}

@@

If we do not restrict our attention to only Horn clauses then there is
no problem  presenting a focused proof system for all of
first-order intuitionistic logic using the connectives $\forall$ and
$\supset$.  The following rules extend those given earlier.
\begin{metanote}
  Why repeat the previous rules? am
\end{metanote}
\[
  \infer[decide]{\Pscr\vdash A}{D \in \Pscr\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[init]{\Pscr\Downarrow A\vdash A}{}
\]
\[
  \infer[\forall L]{\Pscr\Downarrow\forall x. D\vdash A}
        {\Pscr\Downarrow D[t/x]\vdash A}
  \qquad
  \infer[\supset L]{\Pscr\Downarrow B\supset D\vdash A}
        {\Pscr\vdash B\quad \Pscr\Downarrow D\vdash A}
  \qquad
  \infer[\supset R]{\Pscr\vdash B\supset D}{\Pscr, B\vdash D}
  \qquad
  \infer[\forall R]{\Pscr\vdash \forall x.D}{\Pscr\vdash D[y/x]}
\]
As before, the syntactic variable $A$ denotes atomic formulas and, as
usual, the $\forall R$ rule has the restriction that $y$ is not free in
its conclusion.

@@

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog CIC init lp
% LocalWords:  Andreoli's basicstyle lprolog sigs Coq's sortedness Ev
% LocalWords:  backchain gref indt indc coq dest unfoldings str fpc
% LocalWords:  evar nat forall ys elpi Qed lc args ljf HOAS unfoldE
% LocalWords:  FPCs Prolog evars untyped
